
```{r echo = FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(patchwork)
dat <- read_csv(file = "ldt_data.csv")
dat <- dat %>%
  mutate(language = factor(
    x = language, # column to translate
    levels = c(1, 2), # values of the original data in preferred order
    labels = c("monolingual", "bilingual") # labels for display
  ))
theme_set(theme_grey())
```

# Chapter 3

## Data formats

To visualise the experimental reaction time and accuracy data using `ggplot` we first need to perform some data wrangling, and it is this step that can cause friction with novice users of R. Traditionally, wide-format data is the preferred or default format. Wide-format data typically has one row of data for each participant with separate columns for each score or variable. Where there are repeated-measures variables, the dependent variable is split across different columns with one measurement for each condition.

The simulated lexical decision data is currently in wide-format (see Table\ \@ref(tab:wide-data)) where each participant's aggregated [^3] reaction time and accuracy for each level of the within-subject variable is split across multiple columns.

[^3]: In this tutorial we have chosen to gloss over the data wrangling that must occur to get from the raw data to these aggregated values. This type of wrangling requires a more extensive tutorial than this paper can provide but, more importantly, it is still possible to use R for data visualisation having done these preparatory steps using existing workflows that newcomers to R are comfortable with. The aim of this paper is to bypass these initial, often problematic steps and focus on tangible outputs that may then encourage further mastery of reproducible methods.

```{r wide-data, echo=FALSE}
head(dat) %>% knitr::kable(digits = 2, caption="Data in wide format.")
```

This format is popular because it is intuitive to read and easy to enter data into as all the data for one participant is contained within a single row. However, for the purposes of analysis, and particularly for analysis using R, this format is unsuitable because whilst it is intuitive to read by a human, the same is not true for a computer . Wide-format data combines multiple variables in a single column, for example in Table\ \@ref(tab:wide-data), `rt_word` contains information related to both a DV and the level of an IV. Wickham [-@wickham2014tidy] provides a comprehensive overview of the benefits of tidy data as a standard way of mapping a dataset to its structure, but for the purposes of this tutorial there are two important rules: each column should be a *variable* and each row should be an *observation.*

Moving from using wide-form to long-form datasets can require a conceptual shift on the part of the researcher and one that usually only comes with practice and repeated exposure[^4]. For our example dataset, adhering to these rules would produce Table\ \@ref(tab:long). Rather than different observations of the same dependent variable being split across columns, there is now a single column for the DV reaction time, and a single column for the DV accuracy. Each participant now has multiple rows of data, one for each observation (i.e., for each participant there will be as many rows as there are levels of the within-subject IV). Although there is some repetition of age and language group, each row is unique when looking at the combination of measures.

[^4]: That is to say, if you are new to R, know that many before you have struggled with this conceptual shift - it does get better, it just takes time and your preferred choice of cursing.

```{r tidying, echo = FALSE}
long <- pivot_longer(dat, 
                     cols = rt_word:acc_nonword, 
                     names_to = c("dv_type", "condition"), 
                     names_sep = "_", 
                     values_to = "dv")

long2 <- pivot_wider(long, names_from = dv_type, values_from = dv)
```

```{r long, echo = FALSE}
head(long2) %>% knitr::kable(digits = 2, caption = "Data in the correct format for visualization.")
```

The benefits and flexibility of this format will hopefully become apparent as we progress through the tutorial, however, a useful rule of thumb when working with data in R for visualisation is that *anything that shares an axis should probably be in the same column*. For example, a simple bar chart of means for the reaction time DV would display the variable `condition` on the x-axis with bars representing both the `word` and `nonword` data, therefore, these data should be in one column and not split.

## Transforming data

We have chosen a 2 x 2 design with two DVs as we anticipate that this is a design many researchers will be familiar with and may also have existing datasets with a similar structure. However, it is worth normalising that trial-and-error is part of the process of learning how to apply these functions to new datasets and structures. Data visualisation can be a useful way to scaffold learning these data transformations because they can provide a concrete visual check as to whether you have done what you intended to do with your data.

### Step 1: `pivot_longer()`

The first step is to use the function `pivot_longer()` to transform the data to long-form. The pivot functions can be easier to show than tell - you may find it a useful exercise to run the below code and compare the newly created object `long` (Table\ \@ref(tab:long-example)) with the original `dat` Table\ \@ref(tab:wide-data) before reading on.

```{r long1}
long <- pivot_longer(data = dat, 
                     cols = rt_word:acc_nonword, 
                     names_sep = "_", 
                     names_to = c("dv_type", "condition"),
                     values_to = "dv")
```


-   As with the other tidyverse functions, the first argument specifies the dataset to use as the base, in this case `dat`. This argument name is often dropped in examples.

-   `cols` specifies all the columns you want to transform. The easiest way to visualise this is to think about which columns would be the same in the new long-form dataset and which will change. If you refer back to Table\ \@ref(tab:long-example) and Table\ \@ref(tab:wide-data), you can see that `id`, `age`, and `language` all remain, it is the columns that contain the measurements of the DVs that change. The colon notation `first_column:last_column` is used to select all variables from the first column specified to the second.  In our code, `cols` specifies that the columns we want to transform are `rt_word` to `acc_nonword`.

-   `names_sep` specifies how to split up the variable name in cases where it has multiple components. This is when taking care to name your variables consistently and meaningfully pays off. Because the word to the left of the separator (`_`) is always the DV type and the word to the right is always the condition of the within-subject IV, it is easy to automatically split the columns.

-   `names_to` specifies the names of the new columns that will be created. There are two: one for the text to the left of the separator, and one for the text to the right of the separator.

-   Finally, `values_to` names the new column that will contain the measurements, in this case we'll call it `dv`. At this point you may find it helpful to go back and compare `dat` and `long` again to see how each argument matches up with the output of the table.

```{r long-example, echo = FALSE}
head(long) %>% knitr::kable(digits = 2, caption = "Data in long format.")
```

### Step 2: `pivot_wider()`

Because we have two DVs, we need to perform an additional step. In the current long-form dataset, the column `dv` contains both reaction time and accuracy measures and keeping in mind the rule of thumb that *anything that shares an axis should probably be in the same column,* this creates a problem because we cannot plot two different units of measurement on the same axis. To fix this we need to use the function `pivot_wider()`. Again, we would encourage you at this point to compare `long` and `dat_long` with the below code to try and map the connections before reading on.

```{r}
dat_long <- pivot_wider(long, 
                        names_from = "dv_type", 
                        values_from = "dv")
```


-   The first argument is again the dataset you wish to work from, in this case `long`. We have removed the argument name `data` in this example.

-   `names_from` acts somewhat like the reverse of `names_to` from `pivot_longer()`. It will take the values from the variable specified and use these as variable names, i.e., in this case, the values of `rt` and `acc` that are currently in the `dv_type` column, and turn these into the column names.

-   Finally, `values_from` specifies the values to fill the new columns with. In this case, the new columns `rt` and `acc` will be filled with the values that were in `dv`. Again, it can be helpful to compare each dataset with the code to see how it aligns.

This final long-form data should looks like Table\ \@ref(tab:long-example).

We have purposefully used a more complex dataset with two DVs for this tutorial. If you are working with a dataset with only one DV, note that only step 1 of this process would be necessary, potentially with the removal of the `names_sep` argument. Finally, be careful not to calculate demographic descriptive statistics from this long-form dataset. Because the process of transformation has introduced some repetition for these variables, the wide-form dataset where 1 row = 1 participant should be used for demographic information.

## Histogram 2

Now that we have the experimental data in the right form, we can begin to create some useful visualizations. First, to demonstrate how code recipes can be reused and adapted, we will create histograms of reaction time and accuracy. The below code uses the same template as before but changes the dataset (`dat_long`), the bin-widths of the histograms, the `x` variable to display (`rt`/`acc`), and the name of the x-axis.

```{r histogram-rt, fig.cap="Histogram of reaction times."}
ggplot(dat_long, aes(x = rt)) +
  geom_histogram(binwidth = 10, fill = "white", color = "black") +
  scale_x_continuous(name = "Reaction time (ms)")
```

```{r histogram-acc, fig.cap="Histogram of accuracy scores."}
ggplot(dat_long, aes(x = acc)) +
  geom_histogram(binwidth = 1, fill = "white", color = "black") +
  scale_x_continuous(name = "Accuracy (0-100)")
```

## Density plots

The layer system makes it easy to create new types of plots by adapting existing recipes. For example, rather than creating a histogram, we can create a smoothed density plot by calling `geom_density()` rather than `geom_histogram()`. The rest of the code remains identical.

```{r density-rt, fig.cap="Density plot of reaction time."}
ggplot(dat_long, aes(x = rt)) +
  geom_density()+
  scale_x_continuous(name = "Reaction time (ms)")
```

### Grouped density plots

Density plots are most useful for comparing the distributions of different groups of data. Because the dataset is now in long format, it makes it easier to map another variable to the plot because each variable is contained within a single column. 

* In addition to mapping `rt` to the x-axis, we specify the `fill` aesthetic to fill the visualisation of each level of the `condition` variable with different colours. 
* As with the x and y-axis scale functions, we can edit the names and labels of our fill aesthetic by adding on another `scale_*_*()` layer.
* Note that the `fill` here is set inside the `aes()` function, which tells ggplot to set the fill differently for each value in the `condition` column. You cannot specify which colour here (e.g., `fill="red"`), like you could when you set `fill` inside the `geom_*()` function before.

```{r density-grouped, fig.cap="Density plot of reaction times grouped by condition."}
ggplot(dat_long, aes(x = rt, fill = condition)) +
  geom_density()+
  scale_x_continuous(name = "Reaction time (ms)") +
  scale_fill_discrete(name = "Condition",
                      labels = c("Word", "Non-word"))
```

## Scatterplots

Scatterplots are created by calling `geom_point()` and require both an `x` and `y` variable to be specified in the mapping.

```{r point-plot, fig.cap="Point plot of reaction time versus age."}
ggplot(dat_long, aes(x = rt, y = age)) +
  geom_point()
```

A line of best fit can be added with an additional layer that calls the function `geom_smooth()`. The default is to draw a LOESS or curved regression line, however, a linear line of best fit can be specified using `method = "lm"`. By default, `geom_smooth()` will also draw a confidence envelope around the regression line, this can be removed by adding `se = FALSE` to `geom_point()`.

```{r smooth-plot, warning = FALSE, fig.cap="Line of best fit for reaction time versus age."}
ggplot(dat_long, aes(x = rt, y = age)) +
  geom_point() +
  geom_smooth(method = "lm")
```

### Grouped scatterplots

Similar to the density plot, the scatterplot can also be easily adjusted to display grouped data. For `geom_point()`, the grouping variable is mapped to `color` rather than `fill` and the relevant `scale_*_*()` function is added.

```{r scatter-grouped, warning = FALSE,fig.cap="Grouped scatter plot of reaction time versus age by condition."}
ggplot(dat_long, aes(x = rt, y = age, color = condition)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_color_discrete(name = "Condition",
                      labels = c("Word", "Non-word"))
```

## Customisation  2

### Accessible color schemes

One of the drawbacks of using `ggplot` for visualisation is that the default colour scheme is not accessible (or visually appealing). The red and green default palette is difficult for colour-blind people to differentiate, and also does not display well in grey scale. You can specify exact custom colors for your plots, but one easy option is to use the `viridis` scale functions. These take the same arguments as their default sister functions for updating axis names and labels, but display plots in contrasting colors that can be read by color-blind people and that also print well in grey scale. The `viridis` scale functions provide a number of different options for the color -- try setting `option` to any letter from A - E to see the different sets.

```{r viridis, warning = FALSE,fig.cap="Use the viridis color scheme for accessibility."}
ggplot(dat_long, aes(x = rt, y = age, color = condition)) +
  geom_point() +
  geom_smooth(method = "lm") +
  # use "viridis_d" instead of "discrete" for better colors
  scale_color_viridis_d(name = "Condition",
                        labels = c("Word", "Non-word"),
                        option = "E")
```

## Activities 2

Before you move on try the following:

-   Use `fill` to created grouped histograms that display the distributions for `rt` for each `language` group separately and also edit the fill axis labels. Try adding `position = "dodge"` to `geom_histogram()` to see what happens.

`r hide("Solution")`
```{r eval = FALSE}
# fill and axis changes
ggplot(dat_long, aes(x = rt, fill = language)) +
  geom_histogram(binwidth = 10) +
  scale_x_continuous(name = "Reaction time (ms)") +
  scale_fill_discrete(name = "Group",
                      labels = c("Monolingual", "Bilingual"))
# add in dodge
ggplot(dat_long, aes(x = rt, fill = language)) +
  geom_histogram(binwidth = 10, position = "dodge") +
  scale_x_continuous(name = "Reaction time (ms)") +
  scale_fill_discrete(name = "Group",
                      labels = c("Monolingual", "Bilingual"))
```

`r unhide()`

-   Use `scale_*_*()` functions to edit the name of the x and y-axis on the scatterplot

`r hide("Solution")`
```{r warning = FALSE,eval = FALSE}
ggplot(dat_long, aes(x = rt, y = age)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_continuous(name = "Reaction time") +
  scale_y_continuous(name = "Age")
```

`r unhide()`

-   Use `se = FALSE` to remove the confidence envelope from the scatterplots

`r hide("Solution")`
```{r warning = FALSE,eval = FALSE}
ggplot(dat_long, aes(x = rt, y = age)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_x_continuous(name = "Reaction time") +
  scale_y_continuous(name = "Age")
```

-   Remove `method = "lm"` from `geom_smooth()` to produce a curved regression line.

`r hide("Solution")`
```{r warning = FALSE,eval = FALSE}
ggplot(dat_long, aes(x = rt, y = age)) +
  geom_point() +
  geom_smooth() +
  scale_x_continuous(name = "Reaction time") +
  scale_y_continuous(name = "Age")
```

-   Replace the default `scale_fill_*()` on the grouped density plot with the color-blind friendly version.

`r hide("Solution")`
```{r eval = FALSE}
ggplot(dat_long, aes(x = rt, fill = condition)) +
  geom_density()+
  scale_x_continuous(name = "Reaction time (ms)") +
  scale_fill_viridis_d(option = "E",
                      name = "Condition",
                      labels = c("Word", "Non-word"))
```

