```{r echo = FALSE, warning = FALSE, message = FALSE}
library(knitr)
library(tidyverse)
library(patchwork)
dat <- read_csv(file = "stroop_data.csv")
dat <- dat %>%
  mutate(language = factor(
    x = language, # column to translate
    levels = c(1, 2), # values of the original data in preferred order
    labels = c("monolingual", "bilingual") # labels for display
  ))
long <- pivot_longer(dat, 
                     cols = rt_cong:acc_incon, 
                     names_to = c("dv_type", "condition"), 
                     names_sep = "_", 
                     values_to = "dv")

dat_long <- pivot_wider(long, 
                        names_from = "dv_type", 
                        values_from = "dv")
```
# Chapter 5

## Interaction plots

Interaction plots are commonly used to help display or interpret a factorial design. Just like with the bar chart of means, interaction plots represent data summaries and so they are built up with a series of calls to `stat_summary()`.

-   `shape` acts much like `fill` in previous plots, except that rather than producing different colour fills for each level of the IV, the data points are given different shapes.

-   `size` lets you change the size of lines and points. You usually don't want different groups to be different sizes, so this option is set inside the relevant `geom_*()` function, not inside the `aes()` function.

-   `scale_color_manual()` works much like `scale_color_discrete()` except that it lets you specify the colour values manually.

I DON'T KNOW WHY THIS CODE ISN'T LETTING ME ADJUST THE VIRIDIS color OPTION, WHEN I CHANGE IT NOTHING HAPPENS
LISA - it changes for me, but this might be a place to introduce scale_color_manual() ?

```{r ixn-plot, fig.cap="Interaction plot."}
ggplot(dat_long, aes(x = condition, y = rt, 
                     shape = language,
                     group = language,
                     color = language)) +
  stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_summary(fun = "mean", geom = "line") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) +
  scale_color_manual(values = c("blue", "darkorange")) +
  theme_classic()
```

`ggplot` will scale the axis start and end points to the data. If you wish to adjust these, you can use the `scale_` functions, for example, the below code specifies the `limits` (start and end points) of the y-axis should be 0 - 750 ms.

```{r}
ggplot(dat_long, aes(x = condition, y = rt, 
                     group = language,
                     color = language)) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun = "mean", geom = "line") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) +
  scale_color_manual(values = c("blue", "darkorange")) +
  theme_classic() +
  scale_y_continuous(limits = c(0, 750))
```

-   Raincloud plots
-   Heatmaps
-   Saving plots
-   Facets with scatterplot - 
-   p + notation explain what it means, use of layers
- Getting rid of legends

ridge
Alluvial
raincloud
split-violin