
```{r echo = FALSE, warning = FALSE, message = FALSE}
library(knitr)
library(tidyverse)
library(patchwork)
dat <- read_csv(file = "stroop_data.csv")
dat <- dat %>%
  mutate(language = factor(
    x = language, # column to translate
    levels = c(1, 2), # values of the original data in preferred order
    labels = c("monolingual", "bilingual") # labels for display
  ))
long <- pivot_longer(dat, 
                     cols = rt_cong:acc_incon, 
                     names_to = c("dv_type", "condition"), 
                     names_sep = "_", 
                     values_to = "dv")

dat_long <- pivot_wider(long, 
                        names_from = "dv_type", 
                        values_from = "dv")
```
# Chapter 5

## Interaction plots

Interaction plots are commonly used to help display or interpret a factorial design. Just like with the bar chart of means, interaction plots represent data summaries and so they are built up with a series of calls to `stat_summary()`.

-   `shape` acts much like `fill` in previous plots, except that rather than producing different colour fills for each level of the IV, the data points are given different shapes.

-   `size` lets you change the size of lines and points. You usually don't want different groups to be different sizes, so this option is set inside the relevant `geom_*()` function, not inside the `aes()` function.

-   `scale_color_manual()` works much like `scale_color_discrete()` except that it lets you specify the colour values manually. You can specify RGB colour values or a list of predefined colour names - all available options can be found by running `colours()` in the console. Other manual scales are also available, for example, `scale_fill_manual`.

```{r ixn-plot, fig.cap="Interaction plot."}
ggplot(dat_long, aes(x = condition, y = rt, 
                     shape = language,
                     group = language,
                     color = language)) +
  stat_summary(fun = "mean", geom = "point", size = 3) +
  stat_summary(fun = "mean", geom = "line") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) +
  scale_color_manual(values = c("blue", "darkorange")) +
  theme_classic()
``` 

`ggplot` will scale the axis start and end points to the data. If you wish to adjust these, you can use the `scale_` functions, for example, the below code specifies the `limits` (start and end points) of the y-axis should be 0 - 750 ms.

```{r}
ggplot(dat_long, aes(x = condition, y = rt, 
                     group = language,
                     color = language,
                     shape = language)) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun = "mean", geom = "line") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2) +
  scale_color_manual(values = c("blue", "darkorange")) +
  theme_classic() +
  scale_y_continuous(limits = c(0, 750))
```

## Combined interaction plots

A more complex interaction plot can be produced that takes advantage of the layers to visualise not only the  overall interaction, but the change across conditions for each participant.

This code is more complex than all prior code because it does not use a universal mapping of the plot aesthetics. In our code so far, the aesthetic mapping (`aes`) of the plot has been specified in the first line of code as all layers have used the same mapping, however, is is also possible for each layer to use a different mapping. 

* The first call to `ggplot()` sets up the default mappings of the plot,  `x`, `y` and `group` variable.
`geom_point()`  overrides the default mapping by setting its own `colour` to draw the data points from each language group in a different colour. `alpha` is set to a low value to aid readability.
* Similarly, `geom_line()` overrides the default grouping varibale so that a line is drawn to connect the individual data points for each *participant* (`group = id`) rather than each language group, and also sets the colours.
* Finally, the calls to `stat_summary()` remain largely as they were, with the exception of setting `colour = "black"` and `size = 2` so that the overall means and errorbars can be more easily distinguished from the individual data points. 

```{r}
ggplot(dat_long, aes(x = condition, y = rt, group = language, shape = language)) +
  geom_point(aes(colour = language),alpha = .2) +
  geom_line(aes(group = id, colour = language), alpha = .2) +
  stat_summary(fun = "mean", geom = "point", size = 2, colour = "black") +
  stat_summary(fun = "mean", geom = "line", colour = "black") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2, colour = "black") +
  theme_minimal()
```

## Facets

So far we have produced single plots that display all the desired variables in one, however, there are situations in which it may be useful to create separate plots for each level of a variable. The below code is an adaptation of the code used to produce the grouped scatterplot (see Figure X) in which it may be easier to see how the relationship changes when the data are not overlaid. 

* Rather than using `colour = condition` to produce different colours for each level of `condition`, this variable is instead passed to `facet_wrap()`.

```{r}
ggplot(dat_long, aes(x = rt, y = age)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~condition) 
  scale_color_discrete(name = "Condition",
                      labels = c("Congruent", "Incongruent"))
```

As another example, we can use `facet_wrap()` as an alternative to the grouped violin-boxplot (see Figure X) in which the variable `language` is passed to `facet_wrap()` rather than `fill`. 

```{r warning = FALSE}
ggplot(dat_long, aes(x = condition, y= rt)) +
  geom_violin() +
  geom_boxplot(width = .2, fatten = NULL) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .1) +
  facet_wrap(~language) +
  theme_minimal()
```

Finally, note that editing the labels for faceted variables uses a different, and quite frankly altogether opaque and confusing, syntax calling on the `labeller` function.

```{r warning = FALSE}
ggplot(dat_long, aes(x = condition, y= rt)) +
  geom_violin() +
  geom_boxplot(width = .2, fatten = NULL) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .1) +
  facet_wrap(~language, 
             labeller = labeller(language = (c(monolingual = "Monolingual participants", bilingual = "Bilingual participants")))) +
  theme_minimal()
```

## Saving plots

Just like with datasets, plots can be saved to objects. The below code saves the histograms we produced for reaction time and accuracy to objects named `p1` and `p2`. These plots can then be viewed by calling the object name in the console.

```{r}
p1 <- ggplot(dat_long, aes(x = rt)) +
  geom_histogram(binwidth = 10, color = "black")

p2 <- ggplot(dat_long, aes(x = acc)) +
  geom_histogram(binwidth = 1, color = "black") 
```

Importantly, layers can then be added to these saved objects. For example, the below code adds a theme to the plot saved in `p1` and saves it as a new object `p3`. This is important because many of the examples of `ggplot` code you will find in online help forums use the `p +` format to build up plots but fail to explain what this means, which can be confusing to beginners.

```{r}
p3 <- p1 + theme_minimal()
```

## Exporting plots 

In addition to saving plots to objects for further use in R, the function `ggsave()` can be used to save plots to your hard drive. The only required argument for `ggsave` is the file name of the image file you will create, complete with file extension (this can be "eps", "ps", "tex", "pdf", "jpeg", "tiff", "png", "bmp", "svg" or "wmf"). By default, `ggsave()` will save the last plot displayed, however, you can also specify a specific plot object.

```{r eval = FALSE}
ggsave(filename = "my_plot.png") # save last displayed plot
ggsave(filename = "my_plot.png", plot = p3) # save plot p3
```

The width, height and resolution of the image can all be manually adjusted and the help documentation for is useful here (`?ggsave`).

## Multiple plots

As well as creating separate plots for each level of a variable using `facet_wrap()`, you may also wish to display multiple different plots together and the `patchwork` package provides an intuitive way to do this. `patchwork` does not require the use of any functions once it is loaded with `library()`, you simply need to save the plots you wish to combine to objects as above as use the operators `+`, `/` `()` and `|`.

### Combining two plots

Two plots can be combined side-by-side or stacked on top of each other. These combined plots could also be saved to an object and then passed to `ggsave`.

```{r}
p1 + p2 # side-by-side
p1 / p2 # stacked
```

### Combining three or more plots

Three or more plots can be combined in a number of ways and the `patchwork` syntax is relatively easy to grasp with a few examples and a bit of trial and error. First, save the complex interaction plot and faceted violin-boxplot to `p5` and `p6`.

```{r}
p5 <- ggplot(dat_long, aes(x = condition, y = rt, group = language, shape = language)) +
  geom_point(aes(colour = language),alpha = .2) +
  geom_line(aes(group = id, colour = language), alpha = .2) +
  stat_summary(fun = "mean", geom = "point", size = 2, colour = "black") +
  stat_summary(fun = "mean", geom = "line", colour = "black") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .2, colour = "black") +
  theme_minimal()

p6 <- ggplot(dat_long, aes(x = condition, y= rt)) +
  geom_violin() +
  geom_boxplot(width = .2, fatten = NULL) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .1) +
  facet_wrap(~language, 
             labeller = labeller(language = (c(monolingual = "Monolingual participants", bilingual = "Bilingual participants")))) +
  theme_minimal()
```

The exact layout of your plots will depend upon a number of factors. Try running the below examples and adjust the use of the operators to see how they change the layout.

```{r eval = FALSE}
p1 /p5 / p6 
(p1 + p6) / p5 
p6 | p1 / p5 
```

## Customisation part 4

### Axis labels

Previously when we have edited the main axis labels we used the `scale_` functions to do so. These functions are useful to know because they allow you to customise each aspect of the scale, for example, the breaks and limits. However, if you only need to change the main axis label, there is a quicker way to do so using `labs()`. The below code edits the axis labels for the histogram saved in `p1`. The title and subtitle do not conform to APA standards, however, for presentations etc. they can be useful.

```{r}
p5 + labs(x = "Congruency of stimuli",
          y = "Reaction time (ms)",
          title = "Language group by congruency interaction plot",
          subtitle = "Reaction time data shows evidence of bilingual advantage")
```

You can also use `labs()` to remove axis labels, for example, try adjusting the above code to `x = NULL`.

### Non-meaningful colours

So far when we have produced plots with colours, the colours were meaningful in that they represented different levels of a variable, but it is also possible to include colour for purely aesthetic reasons.

The below code adds `fill = language` to the faceted violin-boxplots, in addition to adjusting `alpha` and using the viridis colour palette.

```{r}
ggplot(dat_long, aes(x = condition, y= rt, fill = language)) +
  geom_violin(alpha = .4) +
  geom_boxplot(width = .2, fatten = NULL, alpha = .6) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .1) +
  facet_wrap(~language, 
             labeller = labeller(language = (c(monolingual = "Monolingual participants", bilingual = "Bilingual participants")))) +
  theme_minimal() +
  scale_fill_viridis_d(option = "E")
```

Specifying a `fill` variable has led to R producing a legend for this variable. However, given that the use of colour is not meaningful, this is a waste of plot space. You can remove this legend with the `guides` function.

```{r}
ggplot(dat_long, aes(x = condition, y= rt, fill = language)) +
  geom_violin(alpha = .4) +
  geom_boxplot(width = .2, fatten = NULL, alpha = .6) +
  stat_summary(fun = "mean", geom = "point") +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = .1) +
  facet_wrap(~language, 
             labeller = labeller(language = (c(monolingual = "Monolingual participants", bilingual = "Bilingual participants")))) +
  theme_minimal() +
  scale_fill_viridis_d(option = "E") +
  guides(fill = FALSE)
```

-   Heatmaps


ridge
Alluvial
raincloud
split-violin